A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-01-11 doc/large_strain.tex
Fix some typos and improve some formulations

  \begin{abstract}
    The performance of finite element solvers on modern computer architectures is typically memory bound.
    The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
    In order to improve the performance of iterative solvers within the high-performance computing context, so-called matrix-free methods are
    widely adopted in the fluid mechanics community, where matrix-vector products are computed on-the-fly.

@@

The performance of finite element solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of iterative solvers, so-called matrix-free methods,
where matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
are widely adopted in the fluid mechanics community.
Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017},
{\color{red} which is also limited to the small strain linear elasticity.}
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
%
In order to obtain gradients with respect to the spatial configuration,
{\color{red}
that appear in the tangent operator \eqref{eq:algebraic_tangent} of finite-strain elasticity,}
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

@@

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the referential configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material part of the fourth-order spatial tangent stiffness tensor $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that the operations are typically the same on all elements.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library within the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

@@

{\color{red}
To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
The efficiency of the preconditioner, from the perspective of linear algebra, with the so defined level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library and therefore not considered here.
}

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
At the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
Note that for the matrix-free approach we are limited to smoothers that use only diagonal elements of the operator.
}

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
%
{\color{red}
MPI parallelization is achieved using the standard domain decomposition, where each MPI process ``owns'' only a subset of elements.
To that end, \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library is used.
Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
Same approach is adopted at the multigrid levels.}
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
{\color{red}
Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
}

@@

{\color{red}In this study we limit ourselves to comparison on a node level, which should already indicate the differences between the  matrix-free and matrix-based approaches.}
The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
{\color{red}
We chose to have around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix based approach.
On another hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
}
%
The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
Computations are done on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
{\color{red}
We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
{\color{red}as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless }
the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
{\color{red}
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
{\color{red}so long as the resulting tangent operator is expressed via the \eqref{eq:algebraic_tangent}.}

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for the Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_2d.pdf}
    \caption{2D}
    \label{fig:roofline_2d}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_3d.pdf}
    \caption{3D}
    \label{fig:roofline_3d}
  \end{subfigure}
  \caption{Roofline model for a selection of polynomial degrees. `Peak DP' denotes the peak double precision performance ($P$),
  `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
  and SIMD vectorization are discarded ($P/8$).}%
  \label{fig:roofline}
\end{figure}

{\color{red}
Figure \ref{fig:roofline} shows results of the Roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, the performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

@@

To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
bound, see Figure \ref{fig:roofline}.
Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
}

>>>>

  \begin{abstract}
    The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
    The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
    In order to improve the performance of iterative solvers within the high-performance computing context, so-called matrix-free methods are
    widely adopted in the fluid mechanics community, where matrix-vector products are computed on-the-fly.

@@

The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of iterative solvers, so-called matrix-free methods,
where matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
are widely adopted in the fluid mechanics community.
Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent master's thesis produced by a student of the first two of the authors \cite{Mentler2017},
{\color{red} which is also limited to small strain linear elasticity.omit?}
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements ({\color{red} reference?}).
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
%
In order to obtain gradients with respect to the spatial configuration,
{\color{red}
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem,}
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

@@

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of the displacement field with respect to the referential configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material part of the fourth-order spatial tangent stiffness tensor $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that the operations are typically the same on all elements.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

@@

{\color{red}
To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
The efficiency of the preconditioner with the considered level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
}

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while must default smoothers rely on an explicit  matrix form.}

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
%
{\color{red}
MPI parallelization is achieved using the standard domain decomposition, where each MPI process ``owns'' only a subset of elements.
To that end, \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library is used.
Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
The same approach is adopted on the multigrid levels.}
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
{\color{red}
Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
}

@@

{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
{\color{red}
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
}
%
The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
Computations are done on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
{\color{red}
We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
{\color{red}as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless, }
the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
{\color{red}
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
{\color{red}as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_2d.pdf}
    \caption{2D}
    \label{fig:roofline_2d}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_3d.pdf}
    \caption{3D}
    \label{fig:roofline_3d}
  \end{subfigure}
  \caption{Roofline model for a selection of polynomial degrees. `Peak DP' denotes the peak double precision performance ($P$),
  `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
    and SIMD vectorization are discarded ($P/8$).{\color{red} TODO Improve design}}%
  \label{fig:roofline}
\end{figure}

{\color{red}
Figure \ref{fig:roofline} shows results of the Roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

@@

{\color{red}
To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order elements to two factors.
First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
bound, see Figure \ref{fig:roofline}.
Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
}
